---
title: "Lab 04 Notes"
author: "Fernando Miguez"
date: "`r Sys.Date()`"
fontsize: 12pt
output: pdf_document
bibliography: ../../book/crop-soil.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(BioCro)
library(ggplot2)
library(lattice)
```

# Lab 4: Leaf area index, light geometries and canopy light interception

In this lab we will first explore different approaches to modeling
LAI. First we can model LAI simply as a function of GDD. For example,
[@CliftonBrown2000] modeled Miscanthus LAI as a function of
Thermal Units using 10$^\circ$C as base temperature and a slope of 0.0102.

\begin{figure}[htpb!]
  \centering
  \includegraphics{./figs/CB-LAI.png}
  \caption{Leaf Area Index modeled as a function of Thermal Units accumulation}
  \label{fig:cb-lai}
\end{figure}

To reproduce the code in this lab you would need to install the \texttt{BioCro} R package.

```{r, install-biocro, eval = FALSE}
install.packages("BioCro", repos="http://R-Forge.R-project.org")
library(BioCro)
```

## Implementation of the simple simulation of LAI based on Thermal Time.

```{r, lai-TT}
data(cmiWet)
names(cmiWet) <- c("year","day","solar","avgTemp")
## Solar MJ/m2
## Temp in F
cmi2005 <- cmiWet[cmiWet$year == 2005,]

args(RUEmod)

res <- RUEmod(cmi05$solar, cmi05$avgTemp)

names(res)

res.d <- as.data.frame(res)

ggplot(data = res.d, aes(x = AGDD, y = lai.cum)) + geom_line() + 
  ggtitle("Leaf Area Index modeled as a function of Thermal Units accumulation")
```

The implementation of the model results in Fig \ref{fig:lai-TT}. We
did not limit our estimate of LAI so probably our prediction is too
high (14.9).

A different approach to modeling leaf area index is based on modeling
individual leaves as in [@Lizaso2003]. The model describes three
processes of the life cycle of leaves articulated in a thermal time
framework: expansion, longevity and senescence. Figure
\ref{fig:lizaso} illustrates the idealized progression of leaf
expansion and senescence.

\begin{figure}[htpb!]
  \centering
  \includegraphics{./figs/Lizaso-ind-leaf.png}
  \caption{Leaf Area Index modeled as individual leaves from Lizaso}
  \label{fig:lizaso}
\end{figure}

These type of relationships are normally modeled using a logistic function. In this case:

\[ LA_i = \frac{Ae_i}{1 + \exp(-k e_i (t - te_i)) } \]

where

$Ae_i$ is the final surface area of the $i$th leaf (cm$^2$), $te_i$
the thermal time when the leaf reaches 50% of its final area (growing
degrees after emergence, base temperature 8$^\circ$ C), and ke$_i$ is
a unitless parameter controlling the slope of the curve. 

In this model $Ae_i$ is described using the following relationship

\[ Ae_i = Ae_x \exp(A_1 \dot ((LN_i-LN_x)/(LN_x - 1))^2 + A_2 ((LN_i - LN_x)/(LN_x - 1))^3) \]

$Ae_x$ and $LN_x$ are the area (cm$^2$) and node of the largest leaf
blade, and $LN_i$ the nodal position of the $i$th leaf blade. $A_1$
and $A_2$ are shape parameters; $A_1$ controls the width of the curve
and $A_2$ controls the degree of skewness.

\begin{figure}[htpb!]
  \centering
  \includegraphics{./figs/Lizaso-relationships.png}
  \caption{Leaf relationships for an individual leaf area index model}
  \label{fig:lizaso-relation}
\end{figure}

Testing the model as implemented in \textsf{R} in the package \textsf{BioCro}.

```{r, biocro}
data(weather04)
laiP <- laiParms(lai.method = "ind-leaf-Lizaso")
res <- MaizeGro(weather04, plant.day = 110, emerge.day = 120, 
                harvest.day=300, laiControl = laiP)
```

```{r, maizegro-plot, fig.cap="Individual leaf area model as implemented in R"}
plot(res, plot.kind="LAI")
```

\section{Canopy Architecture}

There are also models for canopy architecture which have considered, in detail, the complex geometries of leaf arrangements, the height of the crop, sun inclination, time of the day, etc.

There are many steps in the description of the light environment in a
canopy. The topic is also described at length in either
[@Thornley2000] and [@Campbell1998]. Since there are
many steps and it is hard to cover all of them in detail I will only
describe a few of them.

Steps in WIMOVAC

\begin{enumerate}
\item Separate solar radiation into direct and diffuse
\item Calculate a canopy extinction coefficient based on sun
  declination and canopy architecture. Diffuse radiation has a
  different extinction coefficient.
\item Calculate the amount of sunlit and shaded leaf area for
  different layers of the canopy
\item Calculate leaf-level photosynthesis and transpiration for each
  layer taking into account the changes in relative humidity, wind,
  light, etc. 
\end{enumerate}

Another motivation for canopy light interception is to understand the
theory behind the derivation of a leaf area index from light
interception data as it is done in the LI-80 Accupar ceptometer.

\begin{figure}[htpb!]
  \centering
  \includegraphics{./figs/earth_geometry.png}
  \caption{Earth geometry}
  \label{fig:earth-geometry}
\end{figure}

\section{Solar radiation beam angles}

Some equations:

\[ \delta = -23.5 \times \cos (\frac{360  (doy + 10)}{365}  dtr) \]

where $dtr$ is degrees to radians $\pi / 180$.

$\delta$ is the angle which the line of the sun's rays at the earth
makes with the equatorial plane and it is called the declination of
the sun. For the Northern Hemisphere, maximum declination varies from
-23.5$^{\circ}$ at the winter solstice on December 22 to
+23.5$^{\circ}$ at the summer solstice on June 22
(Fig. \ref{fig:sun-decl}).

\begin{figure}[htpb!]
  \centering
  \includegraphics{./figs/sun-decl.pdf}
  \caption{Sun declination}
  \label{fig:sun-decl}
\end{figure}

From the time of the day and the day of the year the zenith ($\theta$) angle can
be derived

\[ \cos (\theta) = \sin(\Omega) \cdot \sin(\delta) + \cos(\Omega) \cdot \cos(15 \cdot (t - t_{sn})) \] 

From here our additional assumption is that a certain proportion of
light actually reaches the surface, this is termed atmospheric
transmissivity ($\alpha$). 

\[ I_{dir} = irr \times (\alpha ^{PP_{o}/\cos(\theta)}) \times \cos(\theta) \]

\[ I_{diff} = 0.5 \times irr \times (1 - \alpha ^{PP_{o}/\cos(\theta)}) \times \cos(\theta) \]

where 

\[ irr = \text{Solar irradiance} \]
\[ PP_o = 10,000 / atm.P \]
\[ atm.P = atmospheric pressure \]

\begin{figure}[htpb!]
  \centering
  \includegraphics{./figs/Dir-Diff-Int.pdf}
  \caption{Simulated direct and diffuse light}
  \label{fig:dir-diff-int}
\end{figure}

\subsection{Extinction coefficient}

The extinction coefficient can be modeled as a function of the zenith
angle and canopy architecture. In this case the canopy architecture is
described by one parameter $\chi$ which is the ratio of average
projected area of canopy elements on horizontal and vertical surfaces. 

\[ \chi = \frac{\text{horizontal projection}}{\text{vertical projection}} \]

For a perfect vertical canopy $\chi = 0$ and for a perfect horizontal
canopy $\chi$ approaches infinity.

\begin{figure}[htpb!]
  \centering
  \includegraphics{./figs/chi-crops.png}
  \caption{$\chi$ values for different crops}
  \label{fig:chi-crops}
\end{figure}

\[ k = \frac{\cos(\theta) \sqrt{\chi^2 + \tan{\theta}^2}}{\chi + 1.744 (\chi + 1.183)^{-0.733}} \]

\section{Code}

```{r, sun-declination}
## Calculating the sun declination
del <- function(doy){
  dtr <- pi / 180
  ans <- -23.5 * cos((360 * ( doy + 10) /365) * dtr)
  ans
}

xx <- 1:365
yy <- del(xx)

qplot(xx, yy, geom="line",
       xlab = 'day of the year',
       ylab = 'sun declination')

## Function to calculate the cosine of the zenith angle
coszang <- function(lat, doy, t.d, t.sn = 12){
  
  dtr <- pi/180
  omega <- lat * dtr
  deltR <- del(doy) * dtr
  t.f <- (15 * (t.d - t.sn)) * dtr

  ss <- sin(deltR) * sin(omega)
  cc <- cos(deltR) * cos(omega)

  coszenithangle <- ss + cc * cos(t.f)
  coszenithangle <- ifelse(coszenithangle < 1e-10, 1e-10, coszenithangle) 
  
  coszenithangle
}

lat <- 42
dtr <- pi / 180

k <- 1

doyhr.angle <- expand.grid(hour = 0:23, doy = 1:365, cos.th = NA)

for(i in 1:365){

  for(j in 0:23){

   doyhr.angle[k,3] <- coszang(lat = lat, doy = i, t.d = j)

   k <- k + 1
 }
}

## Include plot

## We can also calculate daylength, sunup and sundown
daylen <- function(lat, doy){

  dtr <- pi/180

  omega <- lat * dtr

  deltR <- del(doy) * dtr
  
  coshour <- -tan(omega) * tan(deltR)
  coshourdeg <- (1/dtr) * coshour
  daylength <- 2 * (1/dtr) * acos(coshour) / 15
  sunup <- 12 - daylength / 2
  sundown <- 12 + daylength / 2

  list(daylength = daylength, sunup = sunup, sundown = sundown)

}

## Testing the daylength function

xx <- 1:365
yy <- daylen(lat = 42, xx)

xyplot(yy$daylength + yy$sunup + yy$sundown ~ xx,
       type = 'l',
       lty = 1:3,
       lwd = 2,
       col = c('blue', 'red', 'purple'),
       xlab = 'Day of the year',
       ylab = 'Hours',
       key=list(text = list(c('day length', 'sun up', 'sun down')),
         lty = 1:3, col = c('blue', 'red', 'purple'),
         lines = TRUE, lwd = 2))

## Proportion of direct and diffuse radiation

light <- function(irr=NULL, lat, doy, t.d, atm.P = 1e5){

  dtr <- pi / 180

  PP.o <- 1e5/atm.P
  Solar_Constant <- 2650
  if(missing(irr)) irr <- Solar_Constant
  alpha <- 0.85

  coszenithangle <- coszang(lat = lat, doy = doy, t.d = t.d)
  
  I.dir <- irr * (alpha^((PP.o)/coszenithangle)) * coszenithangle

  I.diff <- 0.5 * irr * (1 - alpha ^ ((PP.o)/coszenithangle)) * coszenithangle

  list(I.dir = I.dir, I.diff = I.diff)

}


doyhr.light <- expand.grid(hour = 0:23, I.dir = NA, I.diff = NA)

for(j in 0:23){
  tmp <- light(irr = 2000, lat = lat, doy = 190, t.d = j)
  doyhr.light[j + 1,2] <- tmp$I.dir
  doyhr.light[j + 1,3] <- tmp$I.diff
}

## the previous function is vectorized, even if you don't notice
## So there is no need for a loop

ans <- light(irr = 2000, lat = lat, doy = 190, t.d = 0:23)

xyplot(I.dir + I.diff ~ 0:23 , 
       data = ans, type = 'l')

## Illustrating the ellipsoidal distribution

extcoef <- function(psi, chi) {
 ## Enter psi in degrees
 # psi <- psi * (pi/180)
  num <- sqrt(chi^2 + tan(psi)^2) * cos(psi)
  den <- chi + 1.774*(chi + 1.182)^-0.733

  ans <- num/den

  ans

}

xx <- seq(0, 90) 
yy <- extcoef(xx, 1)
```